---
title: "Creating Effective Visualizations with ggplot2"
subtitle: "A modern, accessible introduction using mtcars"
author: "Miguelangel Palma Rojas"
date: last-modified
lang: en
---

## Objective

This notebook shows every analytical step directly. The code prepares the data, checks its structure, and builds four chart types with the explicit `ggplot2` grammar.

## Load packages

```{r}
#| label: packages
#| message: false
#| warning: false
library(dplyr)
library(ggplot2)
library(tibble)
```

## Prepare variables deliberately

`mtcars` stores cylinders, transmission, and gears as numbers. They are categories for this analysis, so the code converts them to factors before plotting.

```{r}
#| label: prepare-data
required_columns <- c("mpg", "cyl", "hp", "wt", "am", "gear")
missing_columns <- setdiff(required_columns, names(mtcars))

if (length(missing_columns) > 0) {
  stop("Missing required columns: ", paste(missing_columns, collapse = ", "))
}

cars <- mtcars |>
  rownames_to_column("model") |>
  mutate(
    cylinders = factor(cyl, levels = c(4, 6, 8)),
    transmission = factor(
      am,
      levels = c(0, 1),
      labels = c("Automatic", "Manual")
    ),
    gears = factor(gear)
  )

glimpse(cars)
```

The prepared dataset contains `r nrow(cars)` vehicles and `r ncol(cars)` variables.

```{r}
#| label: chart-theme
chart_theme <- theme_minimal(base_size = 12) +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

## Bar chart: compare category counts

Use bars when the question is how many observations belong to each group.

```{r}
#| label: cylinder-bars
#| fig-alt: "Bar chart showing the number of cars with four, six, and eight cylinders."
ggplot(cars, aes(x = cylinders, fill = cylinders)) +
  geom_bar(show.legend = FALSE) +
  scale_fill_viridis_d(end = 0.85) +
  labs(
    title = "Cars by cylinder count",
    x = "Cylinders",
    y = "Cars"
  ) +
  chart_theme
```

## Histogram: inspect a continuous distribution

The bin width controls the level of detail and should be chosen explicitly.

```{r}
#| label: horsepower-histogram
#| fig-alt: "Histogram of horsepower values in the mtcars dataset."
ggplot(cars, aes(x = hp)) +
  geom_histogram(
    binwidth = 25,
    boundary = 0,
    fill = "#2A9D8F",
    color = "white"
  ) +
  labs(
    title = "Horsepower distribution",
    x = "Horsepower",
    y = "Cars"
  ) +
  chart_theme
```

## Scatter plot: examine two numeric variables

Color and shape provide redundant category encodings. The fitted line summarizes association, not causality.

```{r}
#| label: efficiency-scatter
#| fig-alt: "Scatter plot of vehicle weight and fuel efficiency, colored by cylinder count and shaped by transmission."
ggplot(
  cars,
  aes(
    x = wt,
    y = mpg,
    color = cylinders,
    shape = transmission
  )
) +
  geom_point(size = 3, alpha = 0.85) +
  geom_smooth(
    aes(group = 1),
    method = "lm",
    se = TRUE,
    color = "#333333"
  ) +
  scale_color_viridis_d(end = 0.85) +
  labs(
    title = "Fuel efficiency falls as vehicle weight rises",
    subtitle = "Point shape identifies transmission type",
    x = "Weight (1,000 lb)",
    y = "Fuel efficiency (mpg)",
    color = "Cylinders",
    shape = "Transmission"
  ) +
  chart_theme
```

## Box plot: compare distributions across groups

The box plot summarizes each distribution while jittered points preserve the small sample.

```{r}
#| label: efficiency-boxplot
#| fig-alt: "Box plots comparing fuel efficiency for automatic and manual transmissions."
ggplot(cars, aes(x = transmission, y = mpg, fill = transmission)) +
  geom_boxplot(width = 0.6, show.legend = FALSE) +
  geom_jitter(width = 0.08, alpha = 0.65) +
  scale_fill_viridis_d(end = 0.85) +
  labs(
    title = "Fuel efficiency by transmission",
    x = "Transmission",
    y = "Fuel efficiency (mpg)"
  ) +
  chart_theme
```

## Conclusion

- The sample is structurally imbalanced: `r sum(cars$cylinders == "8")` of `r nrow(cars)` vehicles have eight cylinders, compared with `r sum(cars$cylinders == "6")` six-cylinder vehicles. This explains why count comparisons should precede distributional comparisons.
- Vehicle weight and fuel efficiency have a Pearson correlation of `r round(cor(cars$wt, cars$mpg), 3)`, indicating a strong negative linear association in this sample.
- Manual-transmission vehicles average `r round(mean(cars$mpg[cars$am == 1]), 1)` mpg versus `r round(mean(cars$mpg[cars$am == 0]), 1)` mpg for automatics, a difference of `r round(mean(cars$mpg[cars$am == 1]) - mean(cars$mpg[cars$am == 0]), 1)` mpg. This is descriptive rather than causal because transmission is confounded with weight, cylinders, and vehicle design.
- The layered charts therefore support a consistent technical finding: heavier, higher-cylinder vehicles dominate the low-efficiency region, while transmission-level differences require multivariable analysis before causal interpretation.

See the [Spanish version](es/index.qmd).
