---
title: "Creating Effective Visualizations with ggplot2"
subtitle: "A modern, accessible introduction using mtcars"
author: "Miguelangel Palma Rojas"
date: last-modified
lang: en
---

```{r}
#| label: setup
#| include: false
source(here::here("R", "analysis.R"))
cars <- prepare_mtcars()
```

## Objective

This notebook rebuilds the original RPubs exercise as a concise guide to the grammar of graphics. Instead of relying on `qplot()`, every chart explicitly defines data, aesthetic mappings, geometric marks, scales, and labels.

The built-in `mtcars` dataset contains `r nrow(cars)` vehicles and `r ncol(cars)` analysis-ready variables after preparation. Because it ships with R, the example is fully reproducible without downloading external data.

## Prepare variables deliberately

Numeric codes such as cylinder count and transmission are categories in this context. Converting them to factors prevents continuous color scales and makes legends meaningful.

```{r}
dplyr::glimpse(cars)
```

## Bar chart: compare category counts

Use bars when the analytical question is “how many observations belong to each group?”

```{r}
#| fig-alt: "Bar chart showing the number of cars with four, six, and eight cylinders."
plot_cylinder_counts(cars, "en")
```

## Histogram: inspect a continuous distribution

A histogram groups a numeric variable into intervals. Bin width is an analytical choice, not a decorative setting.

```{r}
#| fig-alt: "Histogram of horsepower values in the mtcars dataset."
plot_horsepower_distribution(cars, "en")
```

## Scatter plot: examine two numeric variables

The fitted line summarizes the negative association between weight and fuel efficiency. It does not establish causality.

```{r}
#| fig-alt: "Scatter plot of vehicle weight and fuel efficiency, colored by cylinder count and shaped by transmission."
plot_efficiency_weight(cars, "en")
```

## Box plot: compare distributions across groups

Box plots summarize medians, quartiles, and potential outliers. Jittered observations keep the small sample visible.

```{r}
#| fig-alt: "Box plots comparing fuel efficiency for automatic and manual transmissions."
plot_efficiency_boxplot(cars, "en")
```

## Design principles

1. Match the chart to the data types and question.
2. Encode categories as factors before plotting.
3. Use colorblind-friendly scales and redundant encodings when useful.
4. Write titles that state the takeaway.
5. Include units, readable legends, and alternative text.

## Conclusion

Clear graphics come from explicit analytical choices. The same reusable functions generate the [Spanish version](es/index.qmd), ensuring both pages communicate identical results.

